### 2016-05-09 bioc192/intro.tex
editing bioc notes

This book is intended to provide a basic overview of BIOC192 that will have more than enough details to tutor BIOC192 students.
It is highly recommended to listen to the podcasts and add your own notes to this book to get the maximal use out of this book.

@@

\noindent
Riku Takei \\
4$^{th}$ year Biochemistry MSc student\\

>>>>

This book is intended to provide a ``basic'' overview of BIOC192 that will have enough details to tutor BIOC192 students.
It is highly recommended to listen to the podcasts and add your own notes to this book to get the maximal use out of this book.

This book is organised in such a way that the contents/topics of each lecture are more in a logical order (in my opinion), rather than in the order the lecturer presents the material.
I feel like some of the lecturers skip some details, go back to it later on, and go back and forth every now and then.
This  is quite understandable as they only have an hour to go through a certain aspect of the overall concept, and hence after a block of lectures, you usually understand most of what they were saying in the previous lectures.
Since this book is not time limited, I have organised it into  a ``logical order'' as best as I can.

@@

\noindent
Riku Takei \\
5$^{th}$ year Biochemistry MSc student\\

### 2016-05-09 bioc192/lect01.tex
editing bioc notes

Proteins are the workhorses of the cell -- they are required for every process a cell has to do (e.g. DNA replication, apoptosis, etc.)
Proteins are very small (50$\sim$100\AA{}, or 5$\sim$10nm big; a ``nanostructure''), yet they all have unique structure and functions that makes them important for cell survival and growth, and therefore often referred to as ``molecular machines''.

Biochemistry is the study of the molecular details of the biology of the cells/living organisms, including proteins, DNA and RNA.
To understand what components are involved and how they function, you need to know the structure of the thing you're trying to describe -- which, in our case, is the protein.

@@

Amino acids are the building blocks of a protein.
Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).

There are 20 amino acids in our cells that make up the proteins.
All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
Due to the variable side chains, all amino acids except glycine are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.

\includegraphics[width=\textwidth]{aastructure}

@@

As mentioned earlier, all 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.
Cells take advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster (i.e. it gives the protein its functionality, as well as its unique structure).

@@

\includegraphics[width=\textwidth]{20aa}

\section{Charge of Amino Acids}

Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.
Depending on the pH of the solution, there will be more or less protonated molecules.

The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).

When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to become protonated.
Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to become deprotonated.
So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).

Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
However, some amino acid side chains may be charged under physiological condition and contribute to the net charge of the amino acid, and therefore the protein.
This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.

\begin{center}
\includegraphics[width=0.8\textwidth]{aapka}
\end{center}

@@

Some amino acids are modified after translation -- this is called the post-translational modification (PTM).
Types of PTM includes:
\begin{itemize}
    \item Disulfide bond formation between cysteines
    \item Phosphorylation
    \item Glycosylation
    \item Methylation
    \item Adenylation
    \item Iodination
    \item Metal binding
\end{itemize}

@@

\begin{center}
\includegraphics[width=\textwidth]{amidebond}
\end{center}

>>>>

Proteins are the workhorses of the cell -- they are required for every process a cell has to do (e.g. DNA replication, apoptosis, etc.)
Proteins are very small (50$\sim$100\AA{}, or 5$\sim$10nm big; a ``nanostructure''), yet they all have a unique structure and function that makes them important for cell survival and growth, and therefore often referred to as ``molecular machines''.

Biochemistry is the study of the molecular details of the biology of the cells/living organisms, including proteins, DNA and RNA.
\textbf{To understand what components are involved and how they function, you need to know the structure of the thing you're trying to describe}, which, in our case, is the protein.

@@

\textbf{Amino acids are the building blocks of a protein.}
Proteins are made from multiple amino acids joined together as a long polypeptide chain, which in turn folds into a certain structure (see Lecture 2).

There are 20 (major) types of amino acids in our cells that make up the proteins.
All of these 20 amino acids have an amino group (--NH$_3$ group), a carboxyl group (--COOH group), a hydrogen atom, and a variable side chain (--R group) attached to the central carbon atom (C$_{\alpha}$).
Due to the variable side chains, all amino acids (except glycine) are chiral, meaning that they cannot be mirrored (and hence have two forms: D-- and L-- amino acids).
The reason why glycine is the only amino acid that is achiral is because the glycine side chain group is just a hydrogen atom.

% \includegraphics[width=\textwidth]{aastructure}

@@

All 20 amino acids have a common peptide backbone which is made up of the $\alpha$-carbon, carboxyl group, amino group, and a hydrogen atom -- the only thing different between the amino acids are the side chain, or the R group.
\textbf{This means that each of the 20 amino acids have different chemical properties, depending on the side chain group it has.}
The cell takes advantage of this and arrange these amino acids in a certain way so the protein can, for example, carry out chemical reactions faster.
\textbf{In other words, it gives the protein its functionality, as well as its unique structure.}

\section{Charge of Amino Acid Side Chain}

Amino acids have different charges on them, depending on the pH of the environment, as well as the chemical properties of the side chains of the amino acids.
The $\alpha$-amino group ($\alpha$ means that it is joined onto the C$_{\alpha}$) has a pK$_a$ of about 9, whereas the pK$_a$ of the $\alpha$-carboxyl group is about 2 (for all 20 amino acids).
\textbf{When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated.}
Therefore, depending on the pH of the solution, there will be more or less protonated (or deprotonated) molecules in a single amino acid.

The amount of de/protonated amino acids affect the net charge of the protein, and therefore the isoelectric point (pI) of the protein.
\textbf{The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).}

When the pH is low (e.g. 2), the concentration of the hydrogen ion ([H$^+$]) increases, therefore the molecules are likely to be in a \textbf{protonated} state.
Likewise, when the pH is high (e.g. 10), the concentration of the hydrogen ion ([H$^+$]) decreases, therefore the molecules are likely to be in a \textbf{deprotonated} state.

So, in general, when the pH is below the pK$_a$, the molecule is protonated (i.e. H$^+$ ion added), and when the pH is above the pK$_a$, the molecule is deprotonated (i.e. H$^+$ ion removed).

Under physiological condition (pH = 7), the --COOH and the --NH$_3$ groups are charged so the amino acid is uncharged.
However, some amino acid side chains may be charged under physiological condition and this contributes to the net charge of the amino acid, and therefore the charge of the  protein.
This is because different amino acid side chains have different pK$_a$ values, hence different charge at physiological pH.

\begin{center}
% \includegraphics[width=0.8\textwidth]{aapka}
\end{center}

@@

% \includegraphics[width=\textwidth]{20aa}

@@

Some amino acids are modified after the protein has been translated -- this is called the post-translational modification (PTM) of the protein.
Types of PTM includes:
\begin{itemize}
    \item Disulfide bond formation between cysteines
    \item Phosphorylation
    \item Glycosylation
    \item Methylation
    \item Adenylation
    \item Iodination
    \item Metal binding
\end{itemize}

@@

\begin{center}
% \includegraphics[width=\textwidth]{amidebond}
\end{center}

\subsection*{What you should get out of this lecture}

\begin{itemize}
	\item Proteins are made up from amino acids
	\item There are 20 types of amino acids
	\item Each amino acid has a carboxyl group, an amino group, a hydrogen atom, and a variable R group attached to the alpha--carbon (C$_{\alpha}$)
	\item The charge of amino acid side chain can be altered with the pH of the environment
		\begin{itemize}
			\item When pH = pK$_a$, 50\% of the molecules are protonated and 50\% of the molecules are protonated (i.e. ``neutral charge'')
			\item When pH \textless{} pK$_a$, then the side chain is protonated
			\item When pH \textgreater{} pK$_a$, then the side chain is deprotonated
		\end{itemize}
	\item The isoelectric point (pI) is the pH at which the net charge of the protein is 0 (i.e. neutral).
	\item There are four subgroups of amino acids: Non-polar (hydrophobic), polar uncharged, positively charged, and negatively charged amino acids
	\item Post translational modification is when the amino acids are modified after they have been translated into a protein
\end{itemize}

### 2016-05-09 bioc192/lect02.tex
added graphics
editing bioc notes

\section{Bond Rotation}

@@

\subsection{Phi ($\phi$) Angle}

@@

\subsection{Psi ($\Psi$) Angle}

@@

\subsection{Omega ($\omega$) Angle}

mega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).

\subsection{Chi ($\chi$) Angle}

@@

@@

There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.

Specifically speaking, $\alpha$-helix is a type of helix that is present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.

In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g. the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
There is about 3.6 residues per turn, which is about 5.4\AA per turn.
The distance (or height) between 2 residues is about 1.5\AA.
The side chains point outwards from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side). 
$\alpha$-helices are dipole in nature, due to the amide bond polarity, all of which points at a single direction.

There are some residues that disrupt the helical structure, and these are glycine and proline.
Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
This flexibility can cause disruption in the helical structure.
Proline has a unique side chain that connects back to its amino group, forming an imino group, meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.

$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but itcan have more strands).
Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.

Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content.
About 30\% of the residues in a protein is involved in turns.
It is common for the residues to form H-bonds across the turn.
There are many types of turns (\textgreater16), but type I and II are common.

@@

Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
This is not surprising, because if the secondary structures fall in the forbidden region, how did the structure form in the first place?
It can't -- the structure cannot form due to steric hinderance.

This essentially shows that secondary structures have unique phi/psi angle pairs.

>>>>

There are three main secondary structures in a protein: $\alpha$-helix, $\beta$-sheets, and turns.
These secondary structures of a protein has to be made in accordance with the allowed bond angles of the amide plane of the peptide backbone.

\subsection{Bond Rotation}

@@

\subsubsection{Phi ($\phi$) Angle}

@@

\subsubsection{Psi ($\Psi$) Angle}

@@

\subsubsection{Omega ($\omega$) Angle}

Omega angle is the bond angle rotation of the actual peptide bond (N--C bond, NOT N--C$_{\alpha}$ bond).
This bond can only be either 0$^{\circ}$ (\textit{cis}) or 180$^{\circ}$ (\textit{trans}).

\subsubsection{Chi ($\chi$) Angle}

@@

Specifically speaking, $\alpha$-helix is just one of the many types of helices that may be present in a protein -- there are other types of helices (e.g. polyproline helix), but $\alpha$-helix is the most common type.

In $\alpha$-helix, the carbonyl group of the n$^{th}$ residue forms a H-bond to the amino group of the n+4$^{th}$ residue (e.g. the carbonyl group of the 1$^{st}$ residue H-bonds to the amino group of the 5$^{th}$ residue)
There is about 3.6 residues per turn, which is about 5.4\AA{} per turn.
The distance (or height) between 2 residues is about 1.5\AA.
The side chains point \textbf{outwards} from the helix (this provides certain chemical properties on the sides of the helix, e.g. hydrophobic side).
$\alpha$-helices are \textbf{dipole} in nature, due to the amide bond polarity, all of which points in a single direction.

There are some residues that disrupt the helical structure.
These amino acids are glycine and proline.
Glycine side chain is a hydrogen atom, so it causes very little steric hindrance and therefore very flexible.
This flexibility can cause disruption in the helical structure.
Proline has a unique side chain that connects back to its amino group (forming an ``imino group''), meaning that proline introduces a natural `bend' in the polypeptide chain, and thus disrupts the helical structure.

\begin{center}
% \includegraphics[width=0.7\textwidth]{alphahelix}
\end{center}

$\beta$-sheets, or $\beta$-pleated sheets, are formed from H-bond formation between adjacent polypeptide chain (or strands), therefore $\beta$-sheets are formed from 2$\sim$10 strands (but it can have more strands).
Each strand has about 15 amino acids, and the $\beta$-sheets can be either parallel (both strands go from N to C terminal) or anti-parallel (strands go in an opposite direction).
$\beta$-sheets are extended, but pleated (i.e. it's not completely flat), and the side chains point above and below.

\begin{center}
% \includegraphics[width=0.5\textwidth]{betasheet}
% \includegraphics[width=0.4\textwidth]{turns}
\end{center}

Turns are sharp hairpin turns in a protein, and it has a high glycine and proline content (due to its flexibility and unique structure).
About 30\% of the residues in a protein is involved in turns.
It is common for the residues to form H-bonds across the turn.
There are many types of turns (\textgreater16), but type I and II are the most common.

@@

@@

Secondary structures have certain phi/psi angle pairs, and when you overlay these to the plot, it falls into the allowed regions of the plot at a distinct region.
Secondary structures cannot contain any forbidden angle paris, because if they did contain such angle pair(s) there would be a steric hindrance.

This graph essentially shows that secondary structures have unique phi/psi angle pairs.

% \includegraphics[width=\textwidth]{ramachandron}

\subsection*{What you should get out of this lecture}

\begin{itemize}
	\item Primary structure is the amino acid sequence of a protein
		\begin{itemize}
			\item Primary structure/sequence of the protein determines the 3D structure of the protein
		\end{itemize}
	\item Secondary structure is the regular folds/structure that are stabilised by H-bonds between the peptide backbone
	\item Tertiary structure is the 3D structure of a protein and can have biological function(s)
	\item Quaternary structure is a group of tertiary structure combined together as a single protein
	\item Amide bonds of the peptide backbone has a partial double bond characteristic, and therefore cannot be rotated around this bond (amide plane)
	\item Rotation is allowed between the amide planes, but not all angles are allowed due to steric hindrance
	\item Relative to the C$_{\alpha}$ atom, phi $\phi$ angle is the rotation of the amide plane on the amino group side, whereas the psi $\psi$ angle is the rotation of the amide plane on the other side (i.e. the carboxyl side)
	\item Omega $\omega$ angle is the angle of the peptide bond and can only be \textit{cis} or \textit{trans}
	\item Chi $\chi$ angle is the angle of rotation of the side chain group
	\item $\alpha$-helix properties:
		\begin{itemize}
			\item H-bond between the carbonyl group of the n$^{th}$ residue with the amino group of the n+4$^{th}$ residue
			\item 3.6 residues per turn, 5.4\AA{} per turn
			\item 1.5\AA{} between any two amino acid in the helix
			\item Side chains point outwards
			\item Dipole due to the dipole nature of the amide bonds
			\item Glycine and proline are the common helix breakers
		\end{itemize}
	\item $\beta$-helix properties:
		\begin{itemize}
			\item H-bond between adjacent polypeptide chains/strands
			\item Formed from 2$\sim$10 strands
			\item Each strand has about 15 amino acids
			\item It can be parallel or anti-parallel
			\item Side chains points up or down
		\end{itemize}
	\item Properties of a turn:
		\begin{itemize}
			\item About 30\% of residues are involved in turns
			\item proline and glycine rich
			\item There are many types of turns, but types I and II are most common
		\end{itemize}
	\item Ramachandron Plot shows the allowed Phi/Psi angle pairs in a protein
	\item Secondary structures have their own unique phi/psi angle pairs, and therefore have a distinct area on the Ramachandron plot.
\end{itemize}

### 2016-05-09 bioc192/lect03.tex
added graphics
editing bioc notes

@@

Quaternary structure refers to the non-covalent interactions between different domains, which forms a larger protein structure.
The domains are called `subunits', and these assemblies are called dimer (2 subunits), trimers (3 subunits), etc.

@@

@@

>>>>

\begin{center}
% \includegraphics[width=0.7\textwidth]{supersecondary}
\end{center}

@@

Quaternary structure refers to the larger protein structure formed from non-covalent interactions between different domains.
The domains are called `subunits', and these assemblies are called dimer (2 subunits), trimer (3 subunits), etc.

@@

% \includegraphics[width=\textwidth]{stabilisinginteraction}

@@

\begin{center}
% \includegraphics[width=0.4\textwidth]{chaperone}
\end{center}

\subsection*{What you should get out of this lecture}

\begin{itemize}
	\item Protein domains have hydrophobic core and hydrophilic surface, and is stabilised (mainly) by non-covalent interaction
	\item Protein folding is directed predominantly by the hydrophobic forces from the non-polar residues
	\item Protein folding is not a random process
	\item Some proteins require chaperones and/or chaperonnins to fold correctly
	\item Protein misfolding can lead to serious diseases such as Alzheimer's and prion
\end{itemize}

### 2016-05-09 bioc192/lect04.tex
finished lecture 6, onto lecture 7
added graphics
editing bioc notes

@@

@@

@@

Haemoglobin is the tetrameric (4 subunit) form of myoglobin, and it is a blood O$_2$ transporter.
Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.

@@

In contrast, haemoglobin's O$_2$ binding affinity is not that high compared with myoglobin.
As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
This makes sense as haemoglobin has to bind to O$_2$ at the lungs where pO$_2$ is relatively high, but also has to release the O$_2$ molecules at the tissue/muscle, where the pO$_2$ is low.
This allows haemoglobin to be an effective O$_2$ carrier.

@@

@@

\textbf{In sequential model, binding of one substrate changes the conformation of other subunits, which makes it easier for the other subunits to bind the substrate}

@@

>>>>

This lecture part was to give you an idea of how the tertiary and quaternary structures provide the functionality of  a protein, using myoglobin and haemoglobin as an example.

@@

\begin{center}
% \includegraphics[width=0.5\textwidth]{myoglobin}
\end{center}

@@

\begin{center}
% \includegraphics[width=0.4\textwidth]{haem}
\end{center}

@@

Haemoglobin is the tetrameric (4 subunit) form of myoglobin that consists of two $\alpha$ and two $\beta$ subunits, and it is a blood O$_2$ transporter.
Since haemoglobin is a tetramer, it has 4 haem group in total (one on each subunit), and therefore able to carry 4 O$_2$ molecules.
Slight conformational change occurs when the O$_2$ binds to the haemoglobin molecule, which will become important in explaining the cooperative effect of the subunits.

\begin{center}
% \includegraphics[width=0.4\textwidth]{haemoglobin}
\end{center}

@@

In contrast, haemoglobin's O$_2$ binding affinity is not that high compared to myoglobin.
As you can see from the O$_2$ saturation curve, it requires quite a high pO$_2$ for haemoglobin to be saturated with O$_2$.
This makes sense as haemoglobin has to bind to O$_2$ at the lungs where pO$_2$ is relatively high, but also has to release the O$_2$ molecules at the tissue/muscle, where the pO$_2$ is low.
This allows haemoglobin to be an effective O$_2$ carrier.

@@

\begin{center}
% \includegraphics[width=0.4\textwidth]{saturationcurve}
\end{center}

@@

\begin{center}
% \includegraphics[width=\textwidth]{sequential}
\end{center}

\textbf{In sequential model, binding of one substrate changes the conformation of the other subunits, which makes it easier for the other subunits to bind the substrate}

@@

\begin{center}
% \includegraphics[width=0.3\textwidth]{concerted}
\end{center}

\subsection*{What you should get out of this lecture}

\begin{itemize}
	\item Myoglobin:
		\begin{itemize}
			\item Resides in muscles/tissues for O$_2$ storage
			\item Consists of 8 $\alpha$-helices
			\item Haem group located in the middle
			\item Reversible binfing of O$_2$
			\item Hyperbolic O$_2$ binding curve
		\end{itemize}
	\item Haem:
		\begin{itemize}
			\item Four pyrrole rings linked together with Fe$^{2+}$ at the center of the molecule
			\item Fe$^{2+}$ is slightly below the plane of the porphyrin ring when O$_2$ is not bound
			\item Attached to the myoglobin via the His F8
			\item His E7 affects the O$_2$ binding affinity of the haem
		\end{itemize}
	\item Cyanide binds to the Fe$^{3+}$ form, not the Fe$^{2+}$ form
	\item Cyanide's main inhibitory effect is at the cytochrome Fe$^{3+}$, which affects the electron transport system
	\item Haemoglobin:
		\begin{itemize}
			\item Tetrameric form of myglobin
			\item Consists of 2 $\alpha$ and 2 $\beta$ subunits
			\item Blood O$_2$ transporter and carries 4 O$_2$ molecules at a time
			\item Slight conformational change occurs when O$_2$ binds to a subunit
			\item Sigmoidal O$_2$ binding curve
		\end{itemize}
	\item Myoglobin has a higher O$_2$ binding affinity than haemoglobin
	\item Allosterism -- Change in binding affinity caused by a conformational change of the protein at a site that is different from where the substrate binds to
	\item In sequential model, binding of one substrate changes the conformation of the other subunits, which makes it easier for the other subunits to bind the substrate
	\item In concerted model, the T and R states are at equilibrium and substrates can bind at any time to any conformational state. The binding of substrate causes the equilibrium to shift to the R state and the substrates are more likely to bind to the protein
\end{itemize}

### 2016-05-09 bioc192/lect05.tex
added graphics
editing bioc notes

However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
Since these molecules have effects on haemoglobin, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.

\vspace{0.5cm}

@@

@@

@@

>>>>

However, there are other molecules that bind to the haemoglobin \textbf{at a different site as where the O$_2$ binds to}.
Since these molecules have effects on haemoglobin's binding affinity, yet it binds to a distant position of the haemoglobin, these molecules are called the allosteric effectors.

\vspace{0.3cm}

@@

\begin{center}
% \includegraphics[width=0.3\textwidth]{bpg}
\end{center}

@@

\begin{center}
% \includegraphics[width=0.3\textwidth]{satcurve2}
% \includegraphics[width=0.6\textwidth]{satcurveph}
\end{center}

@@

% \includegraphics[width=\textwidth]{bicarbonate}

\subsection*{What you should get out of this lecture}

\begin{itemize}
	\item Allosteric effectors are molecules that bind to the allosteric protein at a place other than/different to the normal substrate binds to.
	\item
	\item
	\item
	\item
	\item
	\item
	\item
	\item
\end{itemize}
